Patient-Specific Neural Field Modeling from Micro-Electrode Data

### 2010-08-03 ltx/data_paper/DataPaper.tex
minor correction
Addition of stochastic method to Datapaper.tex. Note changes in noise covariance matrix.

\ifpdf
\usepackage[pdftex]{graphicx}
\else
\usepackage{graphicx}
\fi
\usepackage{color}
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}

@@

The neural field equations used as the parametric form of the model are based on the influential papers from Wilson and Cowan~\cite{Wilson1973}, and Amari~\cite{Amari1977}. This class of model describes a continuous cortical sheet or surface, relating mean firing rates of pre and post-synaptic neural populations. The neural field, $v\left( {\mathbf{r},t} \right)$, at position $\mathbf{r}$ and time $t$, is the aggregated post-synaptic potentials described by the temporal convolution 

@@

	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right)dt'}, 

@@

where $h(\cdot)$ is the synaptic response kernel and $g(\cdot)$ describes the mean firing rate. Assuming an infinite propagation velocity for action potentials within the field, the incoming firing rate the position $\mathbf{r}$ is described by spatial convolution

@@

	\label{RateBasedInteractions} g\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\textrm{d}\mathbf{r}'}, 

@@

where $v_0$ and $\varsigma$ describe the firing threshold the slope of the sigmoid respectively. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 

@@

	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\textrm{d}\mathbf{r}'dt'.

@@

To arrive at the general integral-differential equation form of the model, we shall express the synaptic response kernel as a Green's function 

@@

	\label{GreensFuncDef} Dh\left( t \right) = \delta \left( t \right), 

@@

where $D$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
	\label{eq:ContinuousWithOperator} 
	Dv\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\textrm{d}\mathbf{r}'}. 
By assuming the synaptic response kernel is of first order has the form of
where $\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function, the operator $D$ is
	D=\frac{d}{dt} + \zeta,
yielding
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\textrm{d}\mathbf{r}'}. 
The neural field equations must be written as a discrete-time finite-dimensional model in order to relate it to patient-specific data. The discrete-time model is found applying the a first-order Euler method giving

@@

	\frac{v\left( \mathbf{r},t+T_s \right)-v\left( \mathbf{r},t \right)}{T_s} + \zeta v\left(\mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right) f\left( {v\left( \mathbf{r}',t \right)}\right)\textrm{d}\mathbf{r}'},

@@

	\label{DiscreteTimeModel} 
	\textrm{d}\mathbf{r}'}, 
where $\xi = 1-T_s\zeta$. Finally, a disturbance term is added, to account for uncertainty and unmodelled inputs, to the IDE form giving

@@

where $e_t(\mathbf{r})$ is $i.i.d.$ such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean spatial Gaussian process with covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}.

@@

	\boldsymbol\Sigma_e \triangleq \mathbf{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top}. 

@@

When taking the derivative of equation~\ref{eq:Voldermont}, the first term described in equation~\ref{eq:def of beta} will disappear. Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
nothing that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \boldsymbol\theta^2}=-2 \Xi_2 
\end{equation}
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.

>>>>

\ifpdf
\usepackage[pdftex]{graphicx}
\else
\usepackage{graphicx}
\fi
\usepackage{color}
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\red}{\textcolor{red}}
\newcommand{\blue}{\textcolor{blue}}

@@

The neural field equations used as the parametric form of the model are based on the influential papers from Wilson and Cowan~\cite{Wilson1973}, and Amari~\cite{Amari1977}. This class of model describes a continuous cortical sheet or surface, relating mean firing rates of pre and post-synaptic neural populations. The neural field, $v\left( {\mathbf{r},t} \right)$, at position $\mathbf{r}$ and time $t$, is the aggregated post-synaptic potentials described by the \red{(temporal convolution - I'd rather say integral equation	)} integral equation 

@@

	\label{SpikesToPotential} v\left( {\mathbf{r},t} \right) = e^{-\zeta t}v\left( {\mathbf{r},0} \right) + \int_{0}^t {h\left( {t - t'} \right)g\left( {\mathbf{r},t'} \right)\textrm{d}t'}, 

@@

where $v\left( {\mathbf{r},0} \right)$ is the membrane voltage at the initial time $t = 0$, $h(\cdot)$ is the synaptic response kernel assumed to be first order and of the form 

@@

$\zeta=\tau^{-1}$, $\tau$ is the synaptic time constant and $\eta(t)$ is the Heaviside step function \blue{(you can keep it here if you wish as it doesn't change the integral equation because of the upper limit. I would remove it)}. $g(\cdot)$ describes the mean firing rate. In practice, the firing rate can rarely be considered as a purely determinstic process because \red{(can someone write down why?)}. We thus choose to introduce a spatially coloured space-time Wiener process $W(\mathbf{r},t)$ to model the flactuations in the mean firing rate and decompose $g(\cdot)$ as
\begin{equation}\label{DecompositionOfg}
g(\mathbf{r},t)\textrm{d}t = \tilde{g}(\mathbf{r},t)\textrm{d}t + \sigma_W \textrm{d}W(\mathbf{r},t) 
where $\sigma_W \ge 0$ is a measure of the introduced space-time randomness. For $\sigma_W = 0$ a pure deterministic process is recovered. As a result of the random component component, for $\sigma_W > 0$, the membrane voltage $v(\mathbf{r},t)$ is itself a stochastic process. Assuming an infinite propagation velocity for action potentials within the field, the deterministic incoming firing rate is described by a spatial convolution \red{citation?}
	\label{RateBasedInteractions} \tilde{g}\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( v\left( \mathbf{r}',t \right) \right)\textrm{d}\mathbf{r}'}, 

@@

where $v_0$ and $\varsigma$ describe the firing threshold the slope of the sigmoid respectively. 
By substituting equation~\ref{DecompositionOfg} into equation~\ref{SpikesToPotential} we obtain the stochastic integral equation model 

@@

	\label{StochasticModel} v\left(\mathbf{r},t\right) = e^{-\zeta t}v\left( {\mathbf{r},0} \right) +
	\int_{0}^t h\left(t - t'\right)\tilde{g}\left(\mathbf{r},t'\right) \textrm{d}t'+ \sigma_W\int_{0}^t h\left(t - t'\right)\textrm{d}W\left(\mathbf{r},t'\right).

@@

 Here we assume that $v\left( {\mathbf{r}},0 \right)$ is independent of the disturbance on the action potentials and assumed to be generated by a known distribution. We stress that $\sigma_W$ does not depend on the field $v({\mathbf{r}},t)$ and hence the noise in equation \ref{StochasticModel} is strictly additive. The general integral-differential equation is given by 
\begin{equation} \label{DifferentialEquationContTime}
 \textrm{d}v(\mathbf{r},t) + \zeta v(\mathbf{r},t)\textrm{d}t = \tilde{g}(\mathbf{r},t)\textrm{d}t + \textrm{d}W(\mathbf{r},t)  ~~~~ t \ge 0, v(\mathbf{r},0) = \red{GP?}
To show that this is indeed the case consider the function $\kappa(v(\mathbf{r},t),t) = v(\mathbf{r},t)e^{\zeta t}$. We note that $\kappa(v(\mathbf{r},t),t)$ is twice differentiable so that we can apply It\hat{o}'s formula to obtain

@@

 \textrm{d}\kappa = e^{\zeta t}g(\mathbf{r},t)\textrm{d}t + e^{\zeta t}\sigma \textrm{d}W(\mathbf{r},t)

@@

\noindent Integrating over $[0,t]$ and multiplying throughout by $e^{-\zeta t}$ gives the required result.
The neural field equations must be written as a discrete-time finite-dimensional model in order to relate it to patient-specific data. The discrete-time model is found applying the a first-order Euler-Maruyama method on equation \ref{DifferentialEquationContTime} giving

@@

	v\left( \mathbf{r},t+T_s \right)-v\left( \mathbf{r},t \right) + \zeta T_sv\left(\mathbf{r},t \right) = T_s\int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right) f\left( {v\left( \mathbf{r}',t \right)}\right)\textrm{d}\mathbf{r}'} + 	\sigma_W[W\left( \mathbf{r},t+T_s \right)-W\left( \mathbf{r},t \right)],

@@

@@

where $e_t(\mathbf{r}) = \sigma_W[W\left( \mathbf{r},t+T_s \right)-W\left( \mathbf{r},t \right)]$ is the increment of a space-time Wiener process, $i.i.d.$ with zero spatial mean such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,T_s\sigma_W^2\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,T_s\sigma_W^2\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean spatial Gaussian process with covariance function $\gamma(\mathbf{r}-\mathbf{r}')$~\cite{Rasmussen2005}.

@@

	\boldsymbol\Sigma_e \triangleq T_s\sigma_W^2\mathbf{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top}d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top}. 

@@

When taking the derivative of equation~\ref{eq:Voldermont}, the first term described in equation~\ref{eq:def of beta} will disappear. Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
noting that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \boldsymbol\theta^2}=-2 \Xi_2 
\end{equation}
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.
